Выпускная квалификационная работа

### 2025-06-02 TeX/CDAM/en_cdam_KhromovGolyandina.tex
Minor fixes

\newcommand{\tX}{\mathsf{X}}
\newcommand{\bfX}{\mathbf{X}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\calT}{\mathcal{T}}
\newcommand{\calH}{\mathcal{H}}
\newcommand{\rmH}{\mathrm{H}}
\newcommand{\iu}{\mathrm{i}}

@@

Singular spectrum analysis (SSA) is one of the methods used for time series analysis~\cite{Golyandina2001}, in which the original time series is transformed into a matrix, called the trajectory matrix, using a given window length $L$. The singular value decomposition (SVD) of this matrix is then analyzed. When the objective is to estimate the signal and its properties from an observed noisy series, the first $r$ components of the SVD are considered, where $r$ is the rank of the signal trajectory matrix. Based on the selected components, the signal estimation is constructed. A distinctive feature of the method is that it does not require the specification of a signal model. However, SSA can also handle a parametric signal model in the form of a sum of products of polynomials, exponentials and sinusoids. The frequency estimation problem plays a special role. The ESPRIT method uses the estimation of the signal subspace based on the $r$ leading left singular vectors of the trajectory matrix SVD to estimate the frequencies present in the signal.
The least squares (LS) version of ESPRIT~\cite{Roy1989} is also known as Hankel SVD (HSVD), and the total least squares (TLS) version is known as HTLS \cite{VanHuffel1994}.
%Пусть сигнал задан в виде суммы синусоид (или комплексных экспонент
% в комплексном случае).
%Свойства оценок частот и оценок сигнала сильно различаются. В
% частности, дисперсия  оценки сигнала имеет порядок $1/N$, в то
% время как дисперсия оценок частот имеет порядок $1/N^3$, где $N$
% --- длина временного ряда и шум белый, гауссовский.

A number of works propose tensor modifications of the SSA and ESPRIT
methods, where the original series is transformed into a tensor,
usually of 3rd order, instead of a matrix \cite{DeLathauwer2011,Papy2009,Trung2024}.  One of the common
variants of tensor decompositions is Higher-Order SVD (HO-SVD), which
generalizes the matrix SVD.

@@

\section{Methods description}
\subsection{Tensor SSA algorithm layout for signal extraction}
The general structure of tensor SSA algorithms based on HO-SVD is as follows (Basic SSA is a special case). Let $\tX$ be the observed object. The tensor dimensions $I$, $L$ and $K$ are considered as the window length; some of these dimensions are expressed in terms of the others, or are fixed. The parameters of the algorithm are the values $R_1$, $R_2$ and $R_3$. These are often chosen to be equal to $r$, but not always.
\begin{enumerate}
  \item
    Embedding to the trajectory tensor: $\bfX = \calT(\tX)$.
  \item
    Tensor decomposition $\bfX =\sum_{i=1}^{I} \sum_{l=1}^{J} \sum_{k=1}^{K}
    \mathcal{Z}_{ilk} U^{(1)}_{i}
    \circ U^{(2)}_{l} \circ U^{(3)}_{k}$.
  \item
    Grouping $\hat\bfX =\sum_{i=1}^{R_1} \sum_{l=1}^{R_2} \sum_{k=1}^{R_3}
    \mathcal{Z}_{ilk} U^{(1)}_{i}
    \circ U^{(2)}_{l} \circ U^{(3)}_{k}$.
  \item
    Obtaining from $\hat\bfX$ the signal estimate $\hat\tX$ based on  the structure of the trajectory tensor and the operation that is inversed  to embedding.
\end{enumerate}

@@

Let $\tX = (\tX^{(1)}, \tX^{(2)}, \ldots, \tX^{(P)})$ be a multi-channel
time series consisting of $P$ single-channel series, also called channels.
\begin{definition}
  The tensor embedding operator for a multi-channel time series with
  window length $L$ (that is, $I=P$, $L$, $K=N-L+1$) such that ${1< L < N}$ is a mapping $\calT_{L}$
  that translates the
  $P$-channel time-series $\tX$ into the tensor $\calX \in
  \mathbb{C}^{L\times K \times P}$ as follows
  $\mathcal{X}_{lkp} =  x_{l+k-1}^{(p)}$, where ${l \in
  \overline{1:L}},\, {k \in \overline{1:K}},\, {p \in \overline{1:P}}$.
\end{definition}

\subsection{Algorithm for signal parameters estimation}
Consider the $P$-channel time series (including the one-channel case $P=1$) with elements
\begin{gather*}
  x_n^{(p)}= \sum_{r=1}^{R} a_r^{(p)} e^{\alpha_r n} e^{\iu\left(2\pi
  \omega_r n + \varphi_r^{(p)}\right)},
\end{gather*}
where the model parameters are the amplitudes $a_r^{(p)} \in
\mathbb{C}\setminus\{0\}$, phases ${\varphi_r^{(p)} \in [0, 2\pi)}$,
the frequencies $\omega_r\in [0, 1/2]$, and the damping factors $\alpha_r \in
\mathbb{R}$. The HO-ESPRIT algorithm that estimates the frequencies
and damping factors of a time series is defined as follows.
After the embedding step the matrix $\mathbf{U} = \mathbf{U}_d =
\left[U_1^{(d)} :
U_2^{(d)}:\ldots : U_{R_d}^{(d)}\right]$ for $d\in \{1, 2,
3\}$ is constructed and the following matrix equation
\begin{equation*}
  \mathbf{U}^{\uparrow}=\mathbf{U}_{\downarrow}\mathbf{Z}
\end{equation*}
is solved with respect to matrix $\mathbf{Z}$, where the up and down
arrows placed
behind the matrix $\mathbf{U}$ stand for deleting its first and last
rows accordingly.
The $R$ largest eigenvalues of the matrix $\mathbf{Z}$ are considered
to be the estimates of the poles $\lambda_r = e^{\alpha_r + 2\pi\iu
\omega_r}$, from which the parameters $\alpha_r$ and $\omega_r$ can be obtained.
%Базовые алгоритмы ESPRIT, использующие траекторную матрицу и SVD,
%можно получить из HO-ESPRIT аналогично тому, как из HO-SSA и HO-MSSA
%можно получить базовые SSA и MSSA.

@@

\section{Comparison of tensor methods with matrix methods}
All numerical comparisons are made using time series that are expressed as sums of sinusoids.

@@

\begin{thebibliography}{10}
  \bibitem{DeLathauwer2011}
  De Lathauwer~L.~(2011). Blind separation of exponential polynomials and the decomposition of a tensor in rank-$(L\_r,L\_r,1)$ terms. {\sl SIAM Journal on Matrix Analysis and Applications}. Vol.~{\bf 32}, Num.~{\bf 4},
  pp.~1451-1474.

@@

@@

  \bibitem{Trung2024}
  Trung Le~T., Abed-Meraim~K., Trung ~N.L.~(2024). Higher-Order Singular Spectrum Analysis for Multichannel Biomedical Signal Analysis. {\sl 2024 32nd European Signal Processing Conference (EUSIPCO), Lyon, France}. Pp.~1337-1341.

  \bibitem{VanHuffel1994}
  Van Huffel~S., Chen~H.,Decanniere~C., van Hecke~P.~(1994). Algorithm for time-domain {NMR} data fitting based on total least squares. {\sl Journal of Magnetic Resonance, Series A}. Vol.~{\bf 110}, Num.~{\bf 2},
  pp.~19-36.

>>>>

\newcommand{\tX}{\mathsf{X}}
\newcommand{\bfX}{\mathbf{X}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\calT}{\mathcal{T}}
\newcommand{\iu}{\mathrm{i}}

@@

Singular spectrum analysis (SSA) is one of the methods used for time
series analysis~\cite{Golyandina2001}, in which the original time
series is transformed into a matrix, called the trajectory matrix,
using a given window length $L$. The singular value decomposition
(SVD) of this matrix is then analyzed. When the objective is to
estimate the signal and its properties from an observed noisy series,
the first $r$ components of the SVD are considered, where $r$ is the
rank of the signal trajectory matrix. Based on the selected
components, the signal estimation is constructed. A distinctive
feature of the method is that it does not require the specification
of a signal model. However, SSA can also handle a parametric signal
model in the form of a sum of products of polynomials, exponentials
and sinusoids. The frequency estimation problem plays a special role.
The ESPRIT method uses the estimation of the signal subspace based on
the $r$ leading left singular vectors of the trajectory matrix SVD to
estimate the frequencies present in the signal.
The least squares (LS) version of ESPRIT~\cite{Roy1989} is also known
as Hankel SVD (HSVD), and the total least squares (TLS) version is
known as HTLS~\cite{VanHuffel1994}.

A number of works propose tensor modifications of the SSA and ESPRIT
methods, where the original series is transformed into a tensor,
usually of 3rd order, instead of a
matrix~\cite{DeLathauwer2011,Papy2005,Trung2024}.  One of the common
variants of tensor decompositions is Higher-Order SVD (HO-SVD), which
generalizes the matrix SVD.

@@

\section{Methods description}
\subsection{Tensor SSA algorithm layout for signal extraction}
The general structure of tensor SSA algorithms based on HO-SVD is as
follows (Basic SSA is a special case). Let $\tX$ be the observed
object. The tensor dimensions $I$, $L$ and $K$ are considered as the
window length; some of these dimensions are expressed in terms of the
others, or are fixed. The parameters of the algorithm are the values
$R_1$, $R_2$ and $R_3$. These are often chosen to be equal to $r$,
but not always.
\begin{enumerate}
  \item
    Embedding to the trajectory tensor $\bfX = \calT(\tX)$.
  \item
    Tensor decomposition $\bfX =\sum_{i=1}^{I} \sum_{l=1}^{J} \sum_{k=1}^{K}
    \mathcal{Z}_{ilk} U^{(1)}_{i}
    \circ U^{(2)}_{l} \circ U^{(3)}_{k}$.
  \item
    Grouping $\hat\bfX =\sum_{i=1}^{R_1} \sum_{l=1}^{R_2} \sum_{k=1}^{R_3}
    \mathcal{Z}_{ilk} U^{(1)}_{i}
    \circ U^{(2)}_{l} \circ U^{(3)}_{k}$.
  \item
    Obtaining from $\hat\bfX$ the signal estimate $\hat\tX$ based on
    the structure of the trajectory tensor and the operation that is
    inversed  to embedding.
\end{enumerate}

@@

Let $\tX = \left(\tX^{(1)}, \tX^{(2)}, \ldots, \tX^{(P)}\right)$ be a
multi-channel
time series consisting of $P$ single-channel series, also called channels.
\begin{definition}
  The tensor embedding operator for a multi-channel time series with
  window length $L$ (that is, $I=P$, $L$, $K=N-L+1$) such that ${1< L
  < N}$ is a mapping $\calT_{L}$
  that translates the
  $P$-channel time-series $\tX$ into the tensor $\calX \in
  \mathbb{C}^{P \times L \times K}$ as follows
  $\mathcal{X}_{plk} =  x_{l+k-1}^{(p)}$, where ${{p \in
    \overline{1:P}},\, l \in
  \overline{1:L}},\, {k \in \overline{1:K}}$.
\end{definition}

\subsection{Algorithm for signal parameters estimation}
Consider the $P$-channel time series (including the single-channel case
$P=1$) with elements
\begin{gather*}
  x_n^{(p)}= \sum_{r=1}^{R} a_r^{(p)} e^{\alpha_r n} e^{\iu\left(2\pi
  \omega_r n + \varphi_r^{(p)}\right)},
\end{gather*}
where the model parameters are the amplitudes $a_r^{(p)} \in
\mathbb{C}\setminus\{0\}$, phases ${\varphi_r^{(p)} \in [0, 2\pi)}$,
the frequencies $\omega_r\in [0, 1/2]$, and the damping factors $\alpha_r \in
\mathbb{R}$. The HO-ESPRIT algorithm that estimates the frequencies
and damping factors of a time series is defined as follows.
After the embedding step the matrix $\mathbf{U} = \mathbf{U}_d =
\left[U_1^{(d)} :
U_2^{(d)}:\ldots : U_{R_d}^{(d)}\right]$ for $d\in \{1, 2,
3\}$ is constructed and the following matrix equation
\begin{equation*}
  \mathbf{U}^{\uparrow}=\mathbf{U}_{\downarrow}\mathbf{Z}
\end{equation*}
is solved with respect to matrix $\mathbf{Z}$, where the up and down
arrows placed
behind the matrix $\mathbf{U}$ stand for deleting its first and last
rows accordingly.
The $R$ largest eigenvalues of the matrix $\mathbf{Z}$ are considered
to be the estimates of the poles $\lambda_r = e^{\alpha_r + 2\pi\iu
\omega_r}$, from which the parameters $\alpha_r$ and $\omega_r$ can be obtained.

@@

\section{Comparison of tensor and matrix methods}
All numerical comparisons are made using time series that are
expressed as sums of sinusoids.

@@

\begin{thebibliography}{10}

@@

  \bibitem{VanHuffel1994}
  Van Huffel~S., Chen~H., Decanniere~C., van Hecke~P.~(1994).
  Algorithm for time-domain {NMR} data fitting based on total least
  squares. {\sl Journal of Magnetic Resonance, Series A}. Vol.~{\bf
  110}, Num.~{\bf 2},
  pp.~19-36.

  \bibitem{DeLathauwer2011}
  De Lathauwer~L.~(2011). Blind separation of exponential polynomials
  and the decomposition of a tensor in rank-$(L\_r,L\_r,1)$ terms.
  {\sl SIAM Journal on Matrix Analysis and Applications}. Vol.~{\bf
  32}, Num.~{\bf 4},
  pp.~1451-1474.

  \bibitem{Trung2024}
  Trung Le~T., Abed-Meraim~K., Trung ~N.L.~(2024). Higher-order
  singular spectrum analysis for multichannel biomedical signal
  analysis. {\sl 2024 32nd European Signal Processing Conference
  (EUSIPCO), Lyon, France}. Pp.~1337-1341.

@@
